Higher-Order unification for free!

### 2024-04-12 paper/main.tex
intro with spellchecker

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
The use of these meta languages facilitates this task in two
key ways. The first and most well know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone of proof construction and proof search.

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a database of ``type-class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
The following three type-class instances state that:
1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
is finite;
2) the predicate \coqIn{nfact n nf}, relating a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent Given this database a type-class solver is expected to
prove the following statement automatically:

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second order parameter \coqIn{P} that stands for
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
called pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.

In order to implement such a search in Elpi we shall describe the encoding
of CIC terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
terms, called \elpiIn{tm}, that features (among others) the following
constructors:

@@

\noindent
Following the standard syntax of $\lambda$Prolog~\cite{miller_nadathur_2012}
the meta level binding of a variable \elpiIn{x} in an expression
\elpiIn{e} is written <<\elpiIn{x\ e}>>, while square brackets delimit a
list of terms separated by comma. For example the term
<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:

@@

\noindent
We now illustrate the encoding of the three instances above as higher-order
logic-programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises and
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
Since \elpiIn{Pm} is an higher-order unification variable
of type \elpiIn{tm -> tm},
with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa}
admits one solution:

@@

\noindent
Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g}
the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
back to the domain \elpiIn{tm} of Coq terms, obtaining the expected solution
\ref{solution:intro}:

@@

% \noindent
In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
the following unification problem has no solution:

@@

\noindent
The root cause of the problems we sketched in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, that in turns makes the 
unification procedures of the meta language weak.
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of CIC is much richer, for efficiency and predictability
reasons automatic proof search procedure typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs some care in order to
simulate a logic program in CIC using the unification of Elpi.

>>>>

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012}, and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
The use of these meta languages facilitates this task in two
key ways. The first and most well-know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone of proof construction and proof search.

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC), and we aim to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based on proof search procedures
reminiscent of Prolog, which back-chain lemmas taken
from a database of ``type-class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta-programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper, we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} identifies types whose inhabitants can be enumerated in a (finite) list.
The following three type-class instances state that:
1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
is finite;
2) the predicate \coqIn{nfact n nf}, relating a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent Given this database, a type-class solver is expected to
prove the following statement automatically:

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.

In order to implement such a search in Elpi, we shall describe the encoding
of CIC terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
terms, called \elpiIn{tm}, that includes (among others) the following
constructors:

@@

\noindent
Following the standard syntax of $\lambda$Prolog~\cite{miller_nadathur_2012},
the meta-level binding of a variable \elpiIn{x} in an expression
\elpiIn{e} is written as <<\elpiIn{x\ e}>>, while square brackets delimit a
list of terms separated by comma. For example, the term
<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:

@@

\noindent
We now illustrate the encoding of the three instances above as higher-order
logic programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises, and
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
Since \elpiIn{Pm} is a higher-order unification variable
of type \elpiIn{tm -> tm},
with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa}
admits one solution:

@@

\noindent
Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g},
the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
back to the domain \elpiIn{tm} of Coq terms, obtaining the expected solution
\ref{solution:intro}:

@@

% \noindent
In turn, this redex prevents rule \ref{clause:r2} from backchaining properly since
the following unification problem has no solution:

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of CIC is much richer, for efficiency and predictability
reasons, automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a logic program in CIC using the unification of Elpi.
